---
title: "Lab2"
author: "Hunter DeVoe"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(pacman, tidyverse, lattice)
```

# Tasks  

## Task 1

* Obtain the working directory by issuing the command getwd():

```{r}
getwd()
```

## Task 2

* Read in the data and print the first six lines

```{r}
epagas <- read.csv("EPAGAS-1.csv", header = TRUE)
head(epagas, 6)
```

## Task 3

* Make the object mpg, the number of miles per gallon vector.

```{r}
mpg <- epagas[['MPG']]
head(mpg, 6)
```


* Transform the mpg variable to z and verify these results.

```{r}
# Now convert to the Z-scale and print contents
z = ((mpg - mean(mpg)) / sd(mpg))
z
```

1. Mean

```{r}
round(mean(z), digits = 4)
```

2. Standard Deviation

```{r}
(sd(z))^2
```

* Using z, find the values of mpg that are possible outliers.


```{r}
mpg[abs(z) >= 2 & abs(z) <= 3]
```

* Using z, find the values of mpg that are defined as outliers. Note that by The Empirical Rule, almost all observations in a data set will have a z-score less than 3 in absolute value. Therefore, we can use this to find any possible outliers in the mpg data set.

```{r}
mpg[(abs(z) > 3)]
```

* Using the lattice package, construct a dotplot with colors, Red = outlier, Blue = possible outlier.

```{r}
# Plot the outliers in red and the possible outliers in blue
mycol = ifelse(abs(z)>3, "Red",
        ifelse(abs(z)>=2 &abs(z)<=3,"Blue", "Black"))

#Plot points with different plotting shapes
mypch = ifelse(abs(z)>3, 20,
        ifelse(abs(z)>=2 &abs(z)<=3,21, 22))

dotplot(mpg,col=mycol)
```


## Task 4

* Make a boxplot of the mpg variable.  
  * Make the box black.
  * Put a notch where the median goes
  * Put a title on the graph.
  * Make the plot horizontal.

```{r}
ggplot(data = epagas, aes(x = MPG)) + geom_boxplot(color = "red", fill = "black", notch = TRUE) + ggtitle("MPG Boxplot")
```

* Using Chebyshev's theorem predict the proportion of data within 2 standard deviations of the mean of the mpg data.    

Chebyshev's theorem states that at least $1 - \frac{1}{k^2}$ of the data lies within k standard deviations of the mean. When k = 2, we get $1 - \frac{1}{2^2}$, which simplifies to $\frac{3}{4}$. Therefore, about 3/4 of the data, or 75%, should be within 2 standard deviations of the mean.  

* Use R to calculate the exact proportion within 2 standard deviations of the mean.

```{r}
length(mpg[abs(z)<2])/length(mpg)
```

* Does Chebyshev agree with the data?

I think Chebyshev would agree. His theorem states that at **least** 3/4, or 75%, of the data is within 2 standard deviations of the mean. We can see that 96% of the data lies within 2 standard deviations of the mean, proving Chebyshev's Rule.  

* Now use the Empirical rule, what proportion (according to the rule) of the data should be within 2 standard deviations of the mean?  
According to the Empirical rule, 95% of the data should be within 2 standard deviations of the mean.

* How well does it correspond?  

I think it corresponds well. We found that 96% of the MPG data is within 2 standard deviations of the mean, which is extremely close to what the empirical rule states.  

* Is the Empirical rule valid in this case? Why?  

One of the best ways to see if the Empirical rule is valid is to create a histogram of the data. The Empirical rule holds only if the data set forms a bell or mound shape, so we can analyze the histogram and see what shape it takes on.

```{r}
hist(mpg, col = rainbow(8))
```

We can see that our histogram does have a similar shape to a bell curve. Therefore, it is safe to assume that the Empirical rule is valid for our data. Why? The Empirical rule states that if a data set is has a mound shaped, symmetric distribution, then about 68% of the data is 1 standard deviation away from the mean, 95% is 2 standard deviations away from the mean, and almost all measurements are 3 standard deviations away from the mean. Since the histogram resembles a bell or a mound, we can safely assume that the Empirical rule holds.
